---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  message= FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# HMSr

<!-- badges: start -->
<!-- badges: end -->

**NOTE: MOST OF THE FUNCTION HAVE NO PRACTICAL USE OUTSIDE [HMS](https://www.hms.is/)**

This pacake will include:

 * **Graphical theme** 
 * **Common helper functions**
    + Tidy seasonal adjustment with X13
 * **Tools for getting and handling data from the datawarehouse**
 * **functions to get external data**

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
remotes::install_github("karsfri/HMSr")
```
## Theme

The default template in ggplot2 is readable, but ugly:

```{r}
library(tidyverse)
library(patchwork)
library(lubridate)
library(HMSr)
data("economics")

economics %>%
  filter(date == lubridate::floor_date(date, "year"), year(date) > 2000) %>%
  gather(var, val, -date) %>%
  ggplot(aes(date, val, color = var)) +
  geom_col(width = 0.3) +
  geom_point() +
  labs(
    title = "Lollipops and stuff",
    subtitle = "With HMS colors",
    x = NULL,
    y = NULL,
    caption = "Source: None of your business :P"
  ) +
  facet_wrap(~var, scales = "free_y")


```

The `theme_set_hms()` function sets `theme_hms` as the default theme, along with default colors for some of the more popular geoms.

Notice also the helper function, `label_isl()`.

```{r}
theme_set_hms()

economics %>%
  filter(date == lubridate::floor_date(date, "year"), year(date) > 2000) %>%
  gather(var, val, -date) %>%
  ggplot(aes(date, val, color = var)) +
  geom_col(width = 0.3) +
  geom_point() +
  theme_hms() +
  scale_color_manual(values = palette_dark) +
  scale_y_continuous(labels = label_isl()) +
  labs(
    title = "Lollipops and stuff",
    subtitle = "With HMS colors",
    x = NULL,
    y = NULL,
    caption = "Source: None of your business :P"
  ) +
  facet_wrap(~var, scales = "free_y")
```

### Caption strings

Strengir fyrir algengar heimildir:

```{r}
c(
  cap_hms,
  cap_hagstofa,
  cap_hms_hagstofa,
  cap_thjodskra,
  cap_hms_thjodskra,
  cap_sedlo,
  cap_hms_sedlo
) %>% 
  knitr::kable()
```

## Seasonal adjustment

TBD


## Dates

The lubriyear converts numeric vector containing years to a date variable at January first of the corrisponding year:

```{r}
x <- tibble(year = 2010:2017)

x %>% 
  mutate(timi = lubriyear(year)) %>% 
  knitr::kable()
```

It also has optional argument, month:
```{r}
x %>% 
  mutate(
    month = rep_len(rep(1:12), 8),
    timi = lubriyear(year, month)
    )
```


lubriYYYYMM works the same, but is made specially for the MMMMYY variable from the datawarehouse:

```{r}
lubriYYYYMM("2017M03") %>% 
  knitr::kable()
```

