---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  message= FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# HMSr

<!-- badges: start -->
<!-- badges: end -->

**NOTE: MOST OF THE FUNCTION HAVE NO PRACTICAL USE OUTSIDE [HMS](https://www.hms.is/)**

This pacake will include:

 * **Graphical theme** 
 * **Common helper functions**
    + Tidy seasonal adjustment with X13
 * **Tools for getting and handling data from the datawarehouse**
 * **functions to get external data**

## Installation

**NOTE: MIGHT NOT WORK IN MAC BECAUSE OF THE FOLLOWING CODE IN theme.R**

```{r}
# Load fonts
  # grDevices::windowsFonts(Setimo = windowsFont("Setimo"))
  # grDevices::windowsFonts(SetimoLight = windowsFont("Setimo Light"))
```

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
remotes::install_github("karsfri/HMSr")
```
## Theme

**NOTE:** color palette will improve! Waiting for more colors

The default template in ggplot2 is readable, but ugly:

```{r}
library(tidyverse)
library(patchwork)
library(lubridate)
library(HMSr)
data("economics")

economics %>%
  filter(date == lubridate::floor_date(date, "year"), year(date) > 2000) %>%
  gather(var, val, -date) %>%
  ggplot(aes(date, val, color = var)) +
  geom_col(width = 0.3) +
  geom_point() +
  labs(
    title = "Lollipops and stuff",
    subtitle = "With default colors :(",
    x = NULL,
    y = NULL,
    caption = "Source: None of your business :P"
  ) +
  facet_wrap(~var, scales = "free_y")


```

The `theme_set_hms()` function sets `theme_hms` as the default theme, along with default colors for some of the more popular geoms.

Notice also the helper function, `label_isl()`.

```{r}
theme_set_hms()

economics %>%
  filter(date == lubridate::floor_date(date, "year"), year(date) > 2000) %>%
  gather(var, val, -date) %>%
  ggplot(aes(date, val, color = var)) +
  geom_col(width = 0.3) +
  geom_point() +
  theme_hms() +
  scale_color_manual(values = palette_dark) +
  scale_y_continuous(labels = label_isl()) +
  labs(
    title = "Lollipops and stuff",
    subtitle = "With HMS colors :)",
    x = NULL,
    y = NULL,
    caption = "Source: None of your business :P"
  ) +
  facet_wrap(~var, scales = "free_y")
```

### Palettes

**WAITING FOR PERMANENT PALETTES**

### Scales



### Caption strings

Strings for common sources to put in caption:
  
  ```{r}
c(
  cap_hms,
  cap_hagstofa,
  cap_hms_hagstofa,
  cap_thjodskra,
  cap_hms_thjodskra,
  cap_sedlo,
  cap_hms_sedlo
) %>% 
  knitr::kable()
```

## save functions
The functions `ggsave_png`, `ggsave_svg` and `ggsave_both` are thin wrappers around `ggsave`, with default resolution and size set. There are two commonly used length and width parameters in *HMS* documents, and those are provided as objects and they have the following values in cm:

```{r}
c(
  width_narrow,
  width_wide,
  height_regular,
  height_full
) %>% 
  knitr::kable()
```

## Seasonal adjustment

*NOT READY*

Group of functions that make X13-ARIMA seasonal adjustment easier on the fly.

* `seas_seasonally_adjusted()`
* `seas_trend()`
* `seas_seasonal_components()`
* `seas_irregular()`
* `seas_irregular()`

```{r}
# https://px.hagstofa.is/pxis/pxweb/is/Efnahagur/Efnahagur__utanrikisverslun__1_voruvidskipti__01_voruskipti/UTA06002.px/table/tableViewLayout1/?rxid=8966fb9b-6a92-41b6-8758-129d490bb573
trade <- read_hagstofan_alt("https://px.hagstofa.is:443/pxis/sq/8ad42406-35c1-442b-a556-1498992b56ed")
```

The data is different components of international trade:

```{r}
trade %>% 
  head() %>% 
  knitr::kable()
```


Seasonal adjustment for one group:


```{r}
trade %>%
  mutate(Mánuður = lubriYYYYMM(Mánuður)) %>%
  ################# Magic :) #################################################
  mutate(
    `Útflutningur fob - Árstíðarleiðrétt` = seas_seasonally_adjusted(`Útflutningur fob`, Mánuður)
    ) %>%
  ##############################################################################
  select(Mánuður, `Útflutningur fob`, `Útflutningur fob - Árstíðarleiðrétt` ) %>%
  gather(var, val, -Mánuður) %>%
  ggplot(aes(Mánuður, val, color = var)) +
  geom_line()
```

## Dates

The lubriyear converts numeric vector containing years to a date variable at January first of the corrisponding year:

```{r}
x <- tibble(year = 2010:2017)

x %>% 
  mutate(timi = lubriyear(year)) %>% 
  knitr::kable()
```

It also has optional argument, month:
```{r}
x %>% 
  mutate(
    month = rep_len(rep(1:12), 8),
    timi = lubriyear(year, month)
    )
```


lubriYYYYMM works the same, but is made specially for the MMMMYY variable from the datawarehouse (and some tables from hagstofan.is):

```{r}
lubriYYYYMM("2017M03") %>% 
  knitr::kable()
```

