---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  message= FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# HMSr

<!-- badges: start -->
<!-- badges: end -->

**NOTE: MOST OF THE FUNCTION HAVE NO PRACTICAL USE OUTSIDE [HMS](https://www.hms.is/)**

This pacake will include:

 * **Graphical theme** 
 * **Common helper functions**
    + Tidy seasonal adjustment with X13
 * **Tools for getting and handling data from the datawarehouse**
 * **functions to get external data**

## Installation

**NOTE: MIGHT NOT WORK IN MAC BECAUSE OF THE FOLLOWING CODE IN theme.R**

```{r}
# Load fonts
  # grDevices::windowsFonts(Setimo = windowsFont("Setimo"))
  # grDevices::windowsFonts(SetimoLight = windowsFont("Setimo Light"))
```

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
remotes::install_github("karsfri/HMSr")
```
## Theme

**NOTE:** color palette will improve! Waiting for more colors

The default template in ggplot2 is readable, but ugly:

```{r setup}
library(tidyverse)
library(patchwork)
library(lubridate)
library(HMSr)
data("economics")

economics %>%
  filter(date == lubridate::floor_date(date, "year"), year(date) > 2000) %>%
  gather(var, val, -date) %>%
  ggplot(aes(date, val, color = var)) +
  geom_col(width = 0.3) +
  geom_point() +
  labs(
    title = "Lollipops and stuff",
    subtitle = "With default colors :(",
    x = NULL,
    y = NULL,
    caption = "Source: None of your business :P"
  ) +
  facet_wrap(~var, scales = "free_y")


```

The `theme_set_hms()` function sets `theme_hms` as the default theme, along with default colors for some of the more popular geoms.

Notice also the helper function, `label_isl()`.

```{r}
theme_set_hms()

economics %>%
  filter(date == lubridate::floor_date(date, "year"), year(date) > 2000) %>%
  gather(var, val, -date) %>%
  ggplot(aes(date, val, color = var)) +
  geom_col(width = 0.3) +
  geom_point() +
  theme_hms() +
  scale_color_manual(values = palette_dark) +
  scale_y_continuous(labels = label_isl()) +
  labs(
    title = "Lollipops and stuff",
    subtitle = "With HMS colors :)",
    x = NULL,
    y = NULL,
    caption = "Source: None of your business :P"
  ) +
  facet_wrap(~var, scales = "free_y")
```

```{r}
  economics %>%
    filter(date == floor_date(date, "year"), year(date) > 2000) %>%
    gather(var, val, -date) %>%
    ggplot(aes(date, val, fill = var)) +
    geom_col() +
    theme_hms() +
    scale_color_manual(values = palette_dark) +
    scale_y_continuous(labels = label_isl()) +
    labs(
      title = "Bars and stuff",
      subtitle = "With HMS colors",
      x = NULL,
      y = NULL,
      caption = "Source: None of your business :P"
    ) +
    facet_wrap(~var, scales = "free_y")
```

```{r}
  economics %>%
    filter(date == floor_date(date, "year"), year(date) > 2000) %>%
    gather(var, val, -date) %>%
    ggplot(aes(date, val, fill = var, color = var)) +
    geom_area(alpha = 0.5, color = NA) +
    geom_line() +
    theme_hms() +
    scale_color_manual(values = palette_dark) +
    scale_y_continuous(labels = label_isl()) +
    labs(
      title = "Area and stuff",
      subtitle = "With HMS colors",
      x = NULL,
      y = NULL,
      caption = "Source: None of your business :P"
    ) +
    facet_wrap(~var, scales = "free_y")
```


### Palettes

**WAITING FOR PERMANENT PALETTES**

### Labels

There are four functions to control the labels of the y-axis and the x-axis. They use comma as a decimal mark and point as a grouping mark. Usually they are used inside `scale_y_continous()` eða `scale_x_continous`, en hér eru þau sýnd inní demo_continous sem sýnir bara einn ás.

```{r}
scales::demo_continuous(c(0, 1000000), labels = label_isk()) 
```
```{r}
scales::demo_continuous(c(0, 1000000), labels = label_isl())
```

```{r}
scales::demo_continuous(c(0, 1), labels = label_percent_isl(accuracy = 0.1))
```

### Caption strings

Strings for common sources to put in caption (for use for the caption argument in `labs()`):
  
  ```{r}
c(
  cap_hms,
  cap_hagstofa,
  cap_hms_hagstofa,
  cap_thjodskra,
  cap_hms_thjodskra,
  cap_sedlo,
  cap_hms_sedlo
) %>% 
  knitr::kable()
```

## save functions
The functions `ggsave_png`, `ggsave_svg` and `ggsave_both` are thin wrappers around `ggsave`, with default resolution and size set. There are two commonly used length and width parameters in *HMS* documents, and those are provided as objects and they have the following values in cm:

```{r}
c(
  width_narrow,
  width_wide,
  height_regular,
  height_full
) %>% 
  knitr::kable()
```

## Seasonal adjustment

*NOT READY*

Group of functions that make X13-ARIMA seasonal adjustment easier on the fly.

* `seas_seasonally_adjusted()`
* `seas_trend()`
* `seas_seasonal_components()`
* `seas_irregular()`
* `seas_irregular()`

```{r}
trade <- read_hagstofan_alt("https://px.hagstofa.is:443/pxis/sq/8ad42406-35c1-442b-a556-1498992b56ed")
```

The [data](https://px.hagstofa.is/pxis/pxweb/is/Efnahagur/Efnahagur__utanrikisverslun__1_voruvidskipti__01_voruskipti/UTA06002.px/table/tableViewLayout1/?rxid=8966fb9b-6a92-41b6-8758-129d490bb573) is different components of international trade:

```{r}
trade %>%
  mutate(Mánuður = lubriYYYYMM(Mánuður)) %>%
  ################# Magic :) #################################################
  mutate(
    `Útflutningur fob - Árstíðarleiðrétt` = seas_seasonally_adjusted(`Útflutningur fob`, Mánuður)
    ) %>%
  ##############################################################################
  select(Mánuður, `Útflutningur fob`, `Útflutningur fob - Árstíðarleiðrétt` ) %>%
  gather(var, val, -Mánuður) %>%
  ggplot(aes(Mánuður, val, color = var)) +
  geom_line()
```

the trend component:

```{r}
trade %>%
  mutate(Mánuður = lubriYYYYMM(Mánuður)) %>%
  ################# Magic :) #################################################
  mutate(
    `Útflutningur fob - Trend` = seas_trend(`Útflutningur fob`, Mánuður)
    ) %>%
  ##############################################################################
  select(Mánuður, `Útflutningur fob`,  `Útflutningur fob - Trend` ) %>%
  gather(var, val, -Mánuður) %>%
  ggplot(aes(Mánuður, val, color = var)) +
  geom_line()
```

### Many groups



```{r}
trade_long <- trade %>%
  mutate(Mánuður = lubriYYYYMM(Mánuður)) %>%
  gather(variable, value, -Mánuður) 

trade_long %>% 
  head() %>% 
  knitr::kable()
```

If the data is in a *long* format or *tidy* format, make sure to group the variables in such a way that all the dates are distinct within each group:

```{r}
trade_long %>% 
  group_by(variable) %>% 
  mutate(trend = seas_trend(value, Mánuður)) %>% 
  ggplot(aes(Mánuður, trend, color = variable)) +
  geom_line()
```

Failing to do so will result in an error:

```{r, error=TRUE}
trade_long %>% 
  # group_by(variable) %>% 
  mutate(trend = seas_trend(value, Mánuður)) %>% 
  ggplot(aes(Mánuður, trend, color = variable)) +
  geom_line()
```

It is possible to get all the different components from the `seasonal::seas()` function with `map_seas()` but it's

still *experimental*. It might result in unexpected behaviour if the data structure is different from this example.

```{r}
trade_seas <- trade_long %>% 
  group_by(variable) %>% 
  map_seas(value, Mánuður) 

trade_seas%>% 
  filter(seasonal_component %in% c("trend", "final", "value")) %>% 
  ggplot(aes(Mánuður, seasonal_value, color = seasonal_component)) +
  geom_line() +
  facet_wrap(~variable, scales = "free_y")
  

```



```{r}
trade_seas%>% 
  filter(seasonal_component %in% c("seasonal")) %>% 
  ggplot(aes(Mánuður, seasonal_value, color = seasonal_component)) +
  geom_line() +
  facet_wrap(~variable, scales = "free_y")
```



## Dates

The lubriyear converts numeric vector containing years to a date variable at January first of the corrisponding year:

```{r}
x <- tibble(year = 2010:2017)

x %>% 
  mutate(timi = lubriyear(year)) %>% 
  knitr::kable()
```

It also has optional argument, month:
```{r}
x %>% 
  mutate(
    month = rep_len(rep(1:12), 8),
    timi = lubriyear(year, month)
    )
```


lubriYYYYMM works the same, but is made specially for the MMMMYY variable from the datawarehouse (and some tables from hagstofan.is):

```{r}
lubriYYYYMM("2017M03") %>% 
  knitr::kable()
```

## Automated graphics

Here is an example of how to use a automatically updated plotting functions. The function *yfirverd*, found in *demo_automation.R*, prints out a plot and returns the data used:

```{r}
df <- yfirverd()
```

The function *yfirverd* also accepts group variables as an argument:

```{r}
df_landshluti <- yfirverd(Landshluti)
```

With the argument *eftir* it's possible to add the names of the groups to the title:

```{r}
df3 <- yfirverd(SerbyliFjolbyli, eftir = "tegund húsnæðis")

```


To prevent the plot from printing, set *print_plot = FALSE*. Then the function only returns the data used:

```{r}
yfirverd(print_plot = FALSE, Landshluti) %>% 
  head() %>% 
  knitr::kable()

```

This can be helpful, e.g. when making a custumized plot or make further calculations. It's also possible to use *plot_yfirverd* to get the plot object, but without facets, for better custumizations:

```{r}
yfirverd(Landshluti, print_plot = FALSE) %>% 
  plot_yfirverd(title = "This is a title") +
  facet_wrap(~Landshluti, ncol = 2)
```
or:

```{r}
windowsFonts(Chiller = windowsFont("Chiller"))


yfirverd(Landshluti, print_plot = FALSE) %>% 
  plot_yfirverd(title = "This is a title") +
  facet_wrap(~Landshluti, ncol = 2) +
  scale_fill_viridis_d(option = "C") + 
  theme(
    text = element_text(family = "Chiller"),
    plot.title = element_text(family = "Chiller"),
    plot.subtitle = element_text(family = "Chiller"),
    strip.text = element_text(family = "Chiller"),
    strip.text.x = element_text(family = "Chiller"),
    plot.caption = element_text(family = "Chiller"),
    )
```

You can add filters to the data with the *filters* argument, but the filters need to be enclosed in a *quos* function call for it to work.

```{r}
df3 <- yfirverd(
  SerbyliFjolbyli, 
  eftir = "tegund húsnæðis", 
  filters = quos(
    SerbyliFjolbyli != "Annað", 
    SerbyliFjolbyli != "Sérhæft íbúðarhúsnæði")
  )

```

## Saving the graphs

The *save_both* function from *theme_hms* is a wrapper for *ggsave* and saves the last plot, both as *svg* and as a *png* at a specified size.

```{r}
# ggsave_both("yfir")
```

Finally, the function *yfirverd_get_data* shows the data that the function *yfirverd* gets from the datawarehouse. 
